---
title: "Predicting the manner of exercises - Peer Graded Assignment ML"
author: "G Raman Kumar"
date: "7th-Feb-2024"
output: html_document
---

```{r, setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Introduction
The goal of this assignment is to use data from accelerometers on the belt,
forearm, arm, and dumbell of 6 participants. They were asked to perform barbell
lifts correctly and incorrectly in 5 different ways. And predict the manner in
which they did the exercise. Train and test datasets provided for download
from the links mentioned on the Coursera project page, are mentioned below.

```{r, train_test}
trainUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFilename <- "pml-traing.csv"
testFilename <- "pml-testing.csv"
# downloading and reading the train and test files
download.file(trainUrl, trainFilename)
training <- read.csv(trainFilename)

download.file(testUrl, testFilename)
testing <- read.csv(testFilename)
```

Decided to use only integer or numeric columns for prediction. Assigned training
data set with only numeric class to another data set. Summarising with summary
function showed many columns with more than nineteen thousand NAs. Decided to
drop all variables with more than nineteen thousand NAs.  

```{r, characters_NAs}
# indexes of all the character columns
char <- which(sapply(training, class)=="character")

# dropping all character columns
workingTrain <- training[,-char]

# determining column names with more than 19000 NAs
na_colnames <- names(workingTrain)[colSums(is.na(workingTrain))>19000]

# Finding the indexes of the above NA columns in the workingTrain dataset
col_indexes <- which(!names(workingTrain) %in% na_colnames)

# Dropping all the columns with NAs greater than 19000
workingTrain <- workingTrain[, col_indexes]

# Keeping only those columns in testing which are there in workingTrain to
# keep both the datasets with same variables.
workingTest <- testing[,col_indexes]

# Attaching the outcome variable "classe" to workingTrain dataset
workingTrain$classe <- training$classe

# Attaching the "problem_id" variable to the workingTest dataset which indicates
# the type of test case each observation is.
workingTest$problem_id <- testing$problem_id
```

Working with the **caret** package.

```{r, loading_library}
library(caret)
model1 <- train(classe ~ ., method = "rf", data = workingTrain,
                trControl = trainControl(method = "repeatedcv", number = 5,
                                         repeats = 2))
```